Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models

### 2020-06-03 analysis/paper/paper.Rmd
recoded Chiwawa R. to Chiwawa River for plots. Made some other text revisions. Added 2 photos as figures. re-knit as html and word

---
title: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@biomark.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@biomark.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richie.Carmichael@biomark.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@biomark.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@biomark.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |
  Needs more...

@@

```{r sp-rec-comp}
# pull out estimates of capacity
cap_tbl = spawn_rec_params %>%
  select(Population, form, est, se, cv) %>%
  arrange(Population, form)

@@

# which watersheds to plot?
keep_wtsds = spawn_rec_params %>%
  filter(form == 'QRF',
         !is.na(est)) %>%
  filter(!Population %in% c('Methow R.')) %>%
  pull(Population)

keep_wtsds = c('Catherine Creek',
               'Chiwawa R.',
               'Upper Grande Ronde River',
               'Hayden Creek',
               'Minam River',
               'Tucannon River',
               'Upper Lemhi',
               'Lostine River',
               'South Fork Salmon River')

@@

Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring volume and area were scaled to the wetted area of each site. Within each category, metrics were ranked according to their MIC value (**Figure ??**). Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.). Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.

@@

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012]. We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010. We used the 90th quantile, rather than a higher quantile, to avoid using predictions that are aimed at the very upper tails of observed fish density, where the variability of predictions may be influenced by sample size issues.

@@

Using the QRF model, we predicted habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin. To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from [@Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior CRB, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale. However, for visualization, predictions of areal capacity (fish/m$^2$) were used.

<!-- Drop that last sentence unless we are going to include an example map of QRF capacity estimates -->

@@

Spawner-recruit data from several watersheds within the interior CRB were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival. A series of spawner-recruit functions were then fit to this data including Beverton-Hold, Ricker, and hockey stick. Estimates of capacity were made from each of these spawner-recruit curves and compared with QRF estimates of capacity (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)).

@@

We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m). We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width:depth ratio, channel unit frequency, standard deviation of the wetted depth, braidedness, total fish cover, percent riparian canopy that was big trees, wetted width, observed discharge, percent fines less than 6mm, lower quantile of substrate size (D16), average august temperature and the frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).

@@

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard (**citation??**). However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults. Such datasets are not available everywhere, nor do they necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single season. Or the dataset could be built over time, sampling different locations in different years. If the latter approach is taken, some thought must be given to possible year effects and how to best deal with them.

@@

The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations. Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases. These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need citations??**).

@@

<!-- Try to include images of high capacity / low capacity sites  -->

@@

```{r sr-table}
spawn_rec_params %>%
  select(Population, data) %>%
  unnest(data) %>%
  group_by(Population) %>%
  summarise(`n Yrs` = n_distinct(BroodYear),
            Spawner_type = unique(Spawner_type)) %>%
  select(Population, `n Yrs`, Spawner_type) %>%
  mutate(Spawner_type = recode(Spawner_type,
                               'total spawners' = 'Spawners',
                               'redd counts' = 'Redds')) %>%
  rename(`Adult Data` = Spawner_type) %>%
  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
                               'Parr Surveys',
                               'RST')) %>%
  full_join(spawn_rec_params %>%
              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
              select(Population, form, prnt_val) %>%
              mutate(form = recode(form,
                                   'BevertonHolt' = 'Beverton Holt',
                                   'Hockey' = 'Hockey Stick')) %>%
              mutate(form = factor(form,
                                   levels = c('Beverton Holt',
                                              'Ricker',
                                              'Hockey Stick',
                                              'QRF'))) %>%
              spread(form, prnt_val)) %>%
  filter(Population %in% keep_wtsds) %>%
  # pander(caption = "(\\#tab:sr-table)Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model. Numbers in parentheses are coefficients of variation.")
  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model. Numbers in parentheses are coefficients of variation.",
        booktabs = T)

@@

```{r wtsd-map, fig.cap = "Watersheds with CHaMP habitat data. Watersheds in black also contain paired fish data. Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
knitr::include_graphics('../figures/ChampWtds.svg')

# for pdf
# knitr::include_graphics('../figures/ChampWtds.png')
```

@@

```{r sr-figure, fig.cap = "Spawner-recruit data from seven watersheds. Solid lines are the spawner-recruit curve, dashed lines are the estimated capacity, and shaded polygons depict the 95% confidence intervals of capacity. Blue corresponds to Beverton-Holt models, green to Ricker models, yellow to hockey stick models, and red to QRF estimates. The QRF solid curve is a Beverton-Holt model with the capacity parameter fixed to the QRF estimate of capacity."}

@@

>>>>

---
title: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@biomark.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@biomark.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richie.Carmichael@biomark.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@biomark.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@biomark.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
# csl: "../templates/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
abstract: |
  Needs more...

@@

```{r sp-rec-comp}
# rename one population
spawn_rec_params %<>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River'))

# pull out estimates of capacity
cap_tbl = spawn_rec_params %>%
  select(Population, form, est, se, cv) %>%
  arrange(Population, form)

@@

# which watersheds to plot?
# keep_wtsds = spawn_rec_params %>%
#   filter(form == 'QRF',
#          !is.na(est)) %>%
#   filter(!Population %in% c('Methow R.')) %>%
#   pull(Population)

keep_wtsds = c('Catherine Creek',
               'Chiwawa River',
               'Upper Grande Ronde River',
               'Hayden Creek',
               'Minam River',
               'Tucannon River',
               'Upper Lemhi',
               'Lostine River',
               'South Fork Salmon River')

@@

Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring volume or area were scaled to the wetted area of each site. Within each category, metrics were ranked according to their MIC value (**Figure ??**). Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.). Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.

***Worth putting the whole MIC habitat selection approach in an appendix or supplemental material, so we can include large figures and tables?***

@@

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012]. We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.

@@

<!-- Using the QRF model, we predicted habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin.  -->

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from [@Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior CRB, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

<!-- However, for visualization, predictions of areal capacity (fish/m$^2$) were used.  -->
<!-- The sentence above was dropped, unless we are going to include an example map of QRF capacity estimates -->

@@

Spawner-recruit data from several watersheds within the interior CRB were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival. A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick, using the *FSA* package [@R-FSA] in R. Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.

@@

We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m). We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average august temperature, wetted width:depth ratio, percent fines less than 6mm, total fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).

@@

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999]. However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults. Such datasets are not available everywhere, nor do they necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single season. Or the dataset could be built over time, sampling different locations in different years. If the latter approach is taken, some thought must be given to possible year effects and how to best deal with them.

@@

The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations. Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is less fine sediment, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases. These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need citations??**) [@Mossop2006].

@@

<!-- Try to include images of high capacity / low capacity sites  -->
See Figures \@ref{fig:site-photo-high} and \@ref{fig:site-photo-low}.

@@

```{r sr-table}
spawn_rec_params %>%
  select(Population, data) %>%
  unnest(data) %>%
  group_by(Population) %>%
  summarise(`n Yrs` = n_distinct(BroodYear),
            Spawner_type = unique(Spawner_type)) %>%
  select(Population, `n Yrs`, Spawner_type) %>%
  mutate(Spawner_type = recode(Spawner_type,
                               'total spawners' = 'Spawners',
                               'redd counts' = 'Redds')) %>%
  rename(`Adult Data` = Spawner_type) %>%
  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
                               'Parr Surveys',
                               'RST')) %>%
  full_join(spawn_rec_params %>%
              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
              select(Population, form, prnt_val) %>%
              mutate(form = recode(form,
                                   'BevertonHolt' = 'Beverton Holt',
                                   'Hockey' = 'Hockey Stick')) %>%
              mutate(form = factor(form,
                                   levels = c('Beverton Holt',
                                              'Ricker',
                                              'Hockey Stick',
                                              'QRF'))) %>%
              spread(form, prnt_val)) %>%
  filter(Population %in% keep_wtsds) %>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River')) %>%
  # pander(caption = "(\\#tab:sr-table)Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model. Numbers in parentheses are coefficients of variation.")
  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model. Numbers in parentheses are coefficients of variation.",
        booktabs = T)

@@

```{r wtsd-map, fig.cap = "Watersheds with CHaMP habitat data. Watersheds in black also contain paired fish data. Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
# knitr::include_graphics('../figures/ChampWtds.png')
knitr::include_graphics('../figures/ChampWtds.jpeg')
```

@@

```{r sr-figure, fig.cap = "Spawner-recruit data from nine watersheds. Solid lines are the spawner-recruit curve, dashed lines are the estimated capacity, and shaded polygons depict the 95% confidence intervals of capacity. Red corresponds to Beverton-Holt models, purple to Ricker models, blue to hockey stick models, and green to QRF estimates. The QRF solid curve is a Beverton-Holt model with the capacity parameter fixed to the QRF estimate of capacity."}

@@

```{r site-photo-high, fig.cap = 'Photo of CHaMP site, CBW05583-415923, with predicted high capacity.'}
knitr::include_graphics('../figures/SitePhoto_CBW05583-415923.jpg')
```

```{r site-photo-low, fig.cap = 'Photo of CHaMP site, CBW05583-109994, with predicted low capacity.'}
knitr::include_graphics('../figures/SitePhoto_CBW05583-109994.jpg')
```
